---
title: Optimize cumulative variations of longitudinal biomarkers to advance personalized
  health care
author: "Yushu Wang"
date: "2023-01-12"
output: html_document
---

## Project: Optimize cumulative variations of longitudinal         ####
            biomarkers to advance personalized health care         ####
## Script purpose: Simulations                                     ####
## Date: Jan 12, 2023                                              ####
## Author: Yushu Wang                                              ####
## Division: Division of Biostatistics, MCW                        ####
## Note: this codes were used to illustrate computational gains    ####

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r}
library(ggplot2)
library(lme4)
library(splines)
library(gridExtra)
library(tidyverse)
```

# simulation
## 4.1 linear mixed effect model
### random intercept
```{r}
set.seed(911)
n <- 1000; obs <- 40
beta <- 15
time <- runif(n*obs, 0.1, 4)
ID <- rep(1:n, obs)
beta0 <- 2

#random effect
random_effect <- rnorm(n, 0, 2)
subject_effect <- random_effect[ID]
var(subject_effect); mean(subject_effect)

epsilon <- rnorm(n*obs, 0, 4)

#data generating model
y <- beta0 + beta*time + subject_effect + epsilon
df2 <- data.frame(time = time, ID = ID, y = y)

#model fitting
me.fit <- lmer(y ~ time + (1 | ID), data = df2)
summary(me.fit)
```

### random slope
```{r}
set.seed(911)
n <- 10000; obs <- 40
beta <- 15; beta0 <- 2
time <- runif(n*obs, 0.1, 4)
ID <- rep(1:n, obs)
group <- sample(c(1,2), size = n, replace = T)
group <- group[ID]

#random effect
sigma_0 <- 1.5; sigma_1 <- 1.0
Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1,
                  0.3*sigma_0*sigma_1, sigma_1^2), nrow = 2)
b <- MASS::mvrnorm(n, c(0.1, 0.2), Sigma)
b0 <- b[,1]
b1 <- b[,2]
B0 <- b0[ID]; B1 <- b1[ID]
mean(b0); mean(b1); cov(b)

epsilon <- rnorm(n*obs, 0, 4)

#data generating 
y <- beta0 + beta*group + B0 + B1*time  + epsilon
df3 <- data.frame(ID = ID, time = time, group = group, y = y)

#fitting model
lme.fit <- lmer(y ~ group + (1 + time  | ID), data = df3)
summary(lme.fit)
```

## LME with splines
### spline on fixed effect
```{r}
set.seed(911)
n <- 10000; obs <- 40
beta0 <- 2
time <- runif(n*obs, 0.1, 4)
ID <- rep(1:n, obs)
group <- sample(c(1,2), size = n, replace = T)
group <- group[ID]

#b-spline
degree <- 3
bs_m <- bs(time, Boundary.knots = range(time), degree = degree)

#generate fixed coefficients
library(MASS)
b_m <- c(10, 15, 20)

#random effect
sigma_0 <- 1.5; sigma_1 <- 1.0
Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1, 0.3*sigma_0*sigma_1, sigma_1^2), nrow = 2)
b <- MASS::mvrnorm(n, c(0.1, 0.2), Sigma)
b0 <- b[,1]
b1 <- b[,2]
B0 <- b0[ID]; B1 <- b1[ID]
mean(b0); mean(b1); cov(b)

epsilon <- rnorm(n*obs, 0, 4)

y <- beta0 + bs_m%*%b_m + B0 + B1*time + epsilon
df4 <- data.frame(ID = ID, time = time, group = group, y = y)

#model fitting
lme.bs.fix <- lmer(y ~ bs(time, Boundary.knots = range(time), degree = degree) + (1 + time | ID), data = df4, control=lmerControl(check.conv.grad= .makeCC("ignore",tol = 2e-3, relTol = NULL)))
summary(lme.bs.fix)

```

### spline on random effect (degree=3)
spline on random effect with fixed intercept only
```{r}
set.seed(1234)
n <- 10000; obs <- 40
time <- rweibull(n*obs, 2)
ID <- rep(1:n, obs)
group <- sample(c(1,0), size = n, replace = T)
group <- group[ID]

#generate random effect coefficients
sigma_0 <- 1.5; sigma_1 <- 1.0; sigma_2 <- 2
Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1, 0.3*sigma_0*sigma_2,
                  0.3*sigma_0*sigma_1, sigma_1^2, 0.3*sigma_1*sigma_2,
                  0.3*sigma_0*sigma_2, 0.3*sigma_1*sigma_2, sigma_2^2), nrow = 3)
b <- MASS::mvrnorm(n, c(0.1, 0.2, 0.3), Sigma)
b0 <- b[,1]
b1 <- b[,2]
b2 <- b[,3]
B0 <- b0[ID]; B1 <- b1[ID]; B2 <- b2[ID]
mean(b0); mean(b1); mean(b2); cov(b)

#b-spline
degree <- 3
bs_m <- bs(time, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)

#fix effect coef
beta <- 1.5; beta0 <- 0.2
epsilon <- rnorm(n*obs, 0, 4)

y <-  beta0 + beta*group + B0*bs_m[,1] + B1*bs_m[,2] + B2*bs_m[,3] + epsilon
df.s2 <- data.frame(ID = ID, time = time, group = group, y = y, bs1 = bs_m[,1], bs2 = bs_m[,2], bs3 = bs_m[,3])

#fitting model
lme.bs2 <- lmer(y ~ group + (-1 + bs1 + bs2 + bs3 | ID), data = df.s2, control=lmerControl(check.conv.grad= .makeCC("ignore",tol = 2e-3, relTol = NULL)))
summary(lme.bs2)
```

spline on random effect with random intercept only
```{r}
set.seed(1234)
n <- 10000; obs <- 40
time <- rweibull(n*obs, 2)
ID <- rep(1:n, obs)
group <- sample(c(1,0), size = n, replace = T)
group <- group[ID]

#generate random effect coefficients
sigma_0 <- 1.5; sigma_1 <- 1.0; sigma_2 <- 2; sigma_3 <- 1.5
Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1, 0.3*sigma_0*sigma_2, 0.3*sigma_0*sigma_3,
                  0.3*sigma_0*sigma_1, sigma_1^2, 0.3*sigma_1*sigma_2, 0.3*sigma_1*sigma_3,
                  0.3*sigma_0*sigma_2, 0.3*sigma_1*sigma_2, sigma_2^2, 0.3*sigma_2*sigma_3,
                  0.3*sigma_0*sigma_3, 0.3*sigma_1*sigma_3, 0.3*sigma_2*sigma_3, sigma_3^2), nrow = 4)
b <- MASS::mvrnorm(n, c(0.1, 0.2, 0.3, 0.15), Sigma)
b0 <- b[,1]
b1 <- b[,2]
b2 <- b[,3]
b3 <- b[,4]
B0 <- b0[ID]; B1 <- b1[ID]; B2 <- b2[ID]; B3 <- b3[ID]
mean(b0); mean(b1); mean(b2); mean(b3); cov(b)

#b-spline
degree <- 3
bs_m <- bs(time, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)

#fix effect coef
beta <- 1.5
epsilon <- rnorm(n*obs, 0, 4)

y <-  beta*group + B0 + B1*bs_m[,1] + B2*bs_m[,2] + B3*bs_m[,3] + epsilon
df.s3 <- data.frame(ID = ID, time = time, group = group, y = y, bs1 = bs_m[,1], bs2 = bs_m[,2], bs3 = bs_m[,3])

#fitting model
lme.bs3 <- lmer(y ~ -1 + group + (bs1 + bs2 + bs3 | ID), data = df.s3, control=lmerControl(check.conv.grad= .makeCC("ignore",tol = 2e-3, relTol = NULL)))
summary(lme.bs3)
```

spline on random effect with fixed and random intercepts
```{r}
set.seed(1234)
n <- 10000; obs <- 40
time <- rweibull(n*obs, 2)
ID <- rep(1:n, obs)
group <- sample(c(1,0), size = n, replace = T)
group <- group[ID]

#generate random effect coefficients
sigma_0 <- 1.5; sigma_1 <- 1.0; sigma_2 <- 2; sigma_3 <- 1.5
Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1, 0.3*sigma_0*sigma_2, 0.3*sigma_0*sigma_3,
                  0.3*sigma_0*sigma_1, sigma_1^2, 0.3*sigma_1*sigma_2, 0.3*sigma_1*sigma_3,
                  0.3*sigma_0*sigma_2, 0.3*sigma_1*sigma_2, sigma_2^2, 0.3*sigma_2*sigma_3,
                  0.3*sigma_0*sigma_3, 0.3*sigma_1*sigma_3, 0.3*sigma_2*sigma_3, sigma_3^2), nrow = 4)
b <- MASS::mvrnorm(n, c(0.1, 0.2, 0.3, 0.15), Sigma)
b0 <- b[,1]
b1 <- b[,2]
b2 <- b[,3]
b3 <- b[,4]
B0 <- b0[ID]; B1 <- b1[ID]; B2 <- b2[ID]; B3 <- b3[ID]
mean(b0); mean(b1); mean(b2); mean(b3); cov(b)

#b-spline
degree <- 3
bs_m <- bs(time, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)

#fix effect coef
beta <- 1.5; beta0 <- 0.2
epsilon <- rnorm(n*obs, 0, 4)

y <- beta0 + beta*group + B0 + B1*bs_m[,1] + B2*bs_m[,2] + B3*bs_m[,3] + epsilon
df.s4 <- data.frame(ID = ID, time = time, group = group, y = y, bs1 = bs_m[,1], bs2 = bs_m[,2], bs3 = bs_m[,3])

#fitting model
lme.bs4 <- lmer(y ~ group + (bs1 + bs2 + bs3 | ID), data = df.s4, control=lmerControl(check.conv.grad= .makeCC("ignore",tol = 2e-3, relTol = NULL)))
summary(lme.bs4)

```

spline on random effect without intercept
```{r}
set.seed(1234)
n <- 10000; obs <- 40
beta <- 1.5
time <- rweibull(n*obs, 2)
ID <- rep(1:n, obs)
group <- sample(c(1,0), size = n, replace = T)
group <- group[ID]

#generate random effect coefficients
sigma_0 <- 1.5; sigma_1 <- 1.0; sigma_2 <- 2
Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1, 0.3*sigma_0*sigma_2,
                  0.3*sigma_0*sigma_1, sigma_1^2, 0.3*sigma_1*sigma_2,
                  0.3*sigma_0*sigma_2, 0.3*sigma_1*sigma_2, sigma_2^2), nrow = 3)
b <- MASS::mvrnorm(n, c(0.1, 0.2, 0.3), Sigma)
b0 <- b[,1]
b1 <- b[,2]
b2 <- b[,3]
B0 <- b0[ID]; B1 <- b1[ID]; B2 <- b2[ID]
mean(b0); mean(b1); mean(b2); cov(b)

epsilon <- rnorm(n*obs, 0, 4)

#b-spline
degree <- 3
bs_m <- bs(time, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)

y <-  beta*group + B0*bs_m[,1] + B1*bs_m[,2] + B2*bs_m[,3] + epsilon
df <- data.frame(ID = ID, time = time, group = group, y = y, bs1 = bs_m[,1], bs2 = bs_m[,2], bs3 = bs_m[,3])

#model fitting
lme.bs <- lmer(y ~ -1 + group + (-1 + bs1 + bs2 + bs3 | ID), data = df, control=lmerControl(check.conv.grad= .makeCC("ignore",tol = 2e-3, relTol = NULL)))
summary(lme.bs)
```

## 4.2 Access different splines
degree = 4
```{r}
set.seed(1234)
n <- 10000; obs <- 40
beta <- 1.5
time <- rweibull(n*obs, 2)
ID <- rep(1:n, obs)
group <- sample(c(1,2), size = n, replace = T)
group <- group[ID]

#generate random effect coefficients
sigma_0 <- 1.5; sigma_1 <- 1.0; sigma_2 <- 2; sigma_3 <- 1.5
Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1, 0.3*sigma_0*sigma_2, 0.3*sigma_0*sigma_3,
                  0.3*sigma_0*sigma_1, sigma_1^2, 0.3*sigma_1*sigma_2, 0.3*sigma_1*sigma_3,
                  0.3*sigma_0*sigma_2, 0.3*sigma_1*sigma_2, sigma_2^2, 0.3*sigma_2*sigma_3,
                  0.3*sigma_0*sigma_3, 0.3*sigma_1*sigma_3, 0.3*sigma_2*sigma_3, sigma_3^2), nrow = 4)
b <- MASS::mvrnorm(n, c(0.1, 0.2, 0.3, 0.15), Sigma)
b0 <- b[,1]
b1 <- b[,2]
b2 <- b[,3]
b3 <- b[,4]
B0 <- b0[ID]; B1 <- b1[ID]; B2 <- b2[ID]; B3 <- b3[ID]
mean(b0); mean(b1); mean(b2); mean(b3); cov(b)

epsilon <- rnorm(n*obs, 0, 4)

#b-spline
degree <- 4
bs_m <- bs(time, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)

y <-  beta*group + B0*bs_m[,1] + B1*bs_m[,2] + B2*bs_m[,3] + B3*bs_m[,4] + epsilon
df <- data.frame(ID = ID, time = time, group = group, y = y, bs1 = bs_m[,1], bs2 = bs_m[,2], bs3 = bs_m[,3], bs4 = bs_m[,4])

#model fitting
lme.bs <- lmer(y ~ -1 + group + (-1 + bs1 + bs2 + bs3 + bs4 | ID), data = df, control=lmerControl(check.conv.grad= .makeCC("ignore",tol = 2e-3, relTol = NULL)))
summary(lme.bs)
```

degree = 4, knots = 3
```{r}
set.seed(1234)
n <- 10000; obs <- 40
beta <- 1.5
time <- rweibull(n*obs, 2)
ID <- rep(1:n, obs)
group <- sample(c(1,2), size = n, replace = T)
group <- group[ID]

#generate random effect coefficients
sigma_0 <- 1.5; sigma_1 <- 1.0; sigma_2 <- 2; sigma_3 <- 1.5; sigma_4 <- 2.5
Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1, 0.3*sigma_0*sigma_2, 0.3*sigma_0*sigma_3, 0.3*sigma_0*sigma_4,
                  0.3*sigma_0*sigma_1, sigma_1^2, 0.3*sigma_1*sigma_2, 0.3*sigma_1*sigma_3, 0.3*sigma_1*sigma_4,
                  0.3*sigma_0*sigma_2, 0.3*sigma_1*sigma_2, sigma_2^2, 0.3*sigma_2*sigma_3, 0.3*sigma_2*sigma_4,
                  0.3*sigma_0*sigma_3, 0.3*sigma_1*sigma_3, 0.3*sigma_2*sigma_3, sigma_3^2, 0.3*sigma_3*sigma_4,
                  0.3*sigma_0*sigma_4, 0.3*sigma_1*sigma_4, 0.3*sigma_2*sigma_4, 0.3*sigma_3*sigma_4, sigma_4^2), nrow = 5)
b <- MASS::mvrnorm(n, c(0.1, 0.2, 0.3, 0.2, 0.1), Sigma)
b0 <- b[,1]
b1 <- b[,2]
b2 <- b[,3]
b3 <- b[,4]
b4 <- b[,5]
B0 <- b0[ID]; B1 <- b1[ID]; B2 <- b2[ID]; B3 <- b3[ID]; B4 <- b4[ID]
mean(b0); mean(b1); mean(b2); mean(b3); mean(b4); cov(b)

epsilon <- rnorm(n*obs, 0, 4)

#b-spline
degree <- 4
bs_m <- bs(time, Boundary.knots = quantile(time, c(0,0.75)), knots = median(time), degree = degree)

y <-  beta*group + B0*bs_m[,1] + B1*bs_m[,2] + B2*bs_m[,3] + B3*bs_m[,4] + B4*bs_m[,5] + epsilon
df <- data.frame(ID = ID, time = time, group = group, y = y, bs1 = bs_m[,1], bs2 = bs_m[,2], bs3 = bs_m[,3], bs4 = bs_m[,4], bs5 = bs_m[,5])

#model fitting
lme.bs <- lmer(y ~ -1 + group + (-1 + bs1 + bs2 + bs3 + bs4 + bs5| ID), data = df, control=lmerControl(check.conv.grad= .makeCC("ignore",tol = 2e-3, relTol = NULL)))
summary(lme.bs)
```

Cumulative variation and plots (degree=3, degree=4 with little changes of the codes)
```{r}
#predict data
long_time <- c()
ID <- c()
group <- c()
for (id in c(1:50)) {
  long_time <- append(long_time, seq(min(df$time[df$ID == id]), max(df$time[df$ID == id]), by = max(df$time)/1000))
  ID <- append(ID, rep(id, length(seq(min(df$time[df$ID == id]), max(df$time[df$ID == id]), by = max(df$time)/1000))))
  group <- append(group, rep(df$group[df$ID == id][1], length(seq(min(df$time[df$ID == id]), max(df$time[df$ID == id]), by = max(df$time)/1000))))
}

bs_m_test <- bs(long_time, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)
new_sim <- data.frame(ID = ID, time = long_time, group = group)
new_sim <- cbind(new_sim, bs_m_test)
names(new_sim)[4:6] <- c("bs1", "bs2", "bs3") #, "bs4", "bs5")
new_sim$pred <- predict(lme.bs, new_sim)

#arc length
degree <- 3;interval <- 0.001;fixcoef <- fixef(lme.bs); rancoef <- ranef(lme.bs)$ID #degree = 4
arc <- c()
for (id in c(1:50)) {
  endtime <- max(new_sim$time[new_sim$ID == id])
  starttime <- min(new_sim$time[new_sim$ID == id])
  group_index <- new_sim$group[new_sim$ID == id][1]
  b_vec <- unlist(c(fixcoef, rancoef[rownames(rancoef) == id,]))
  gt <- seq(starttime, endtime, by = interval)
  b_mat <- cbind(group_index, bs(gt, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)) # ,knots = median(time)
  g_grid <- b_mat %*% b_vec
  arc <- append(arc, sum(sqrt((g_grid[-length(gt)]-g_grid[-1])^2+interval^2)))
}


#plot 
sim_plot <- list() #sim_plotd4, sim_plotd4k3
select_sim_id <- c(29, 30, 3, 48, 38, 18)
for (id in c(1:1)) {
  i <- which(select_sim_id == id)
  df1 <- df[df$ID == id,]
  df2 <- new_sim[new_sim$ID == id,]
  
  p <- ggplot() + 
    geom_point(df1, mapping = aes(time, y), size = 1) + 
    geom_line(df2, mapping = aes(time, pred), color = "blue") +
    annotate(geom = "text", x = 2.5, y = 20, label = paste("cumulative variation =", round(arc[id],3)), hjust = 1) +
    xlim(0, 2.5) +
    ylim(-20, 20) +
    #ggtitle(paste0("Subject ", id, " (Group ", df1$group[1]+1,")")) +
    ggtitle("Scenario 1") +
    theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5))
  sim_plot[[i]] <- p
}
grid.arrange(grobs = sim_plot, ncol = 3)
#grid.arrange(grobs = list(sim_plot[[1]], sim_plotd4[[1]], sim_plotd4k3[[1]]), ncol = 3)
```

## 4.3 Bias and MSE
function to generate data
```{r}
beta<-1.5; sigma_0<-1.5; sigma_1<-1; sigma_2<-2
simulation <- function(n, obs, seed=NULL, degree=3){
  if (!is.null(seed)) set.seed(seed)
  time <- rweibull(n*obs, 2)
  ID <- rep(1:n, obs)
  group <- sample(c(1,0), size = n, replace = T)
  group <- group[ID]
  
  #generate random effect coefficients
  Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1, 0.3*sigma_0*sigma_2,
                  0.3*sigma_0*sigma_1, sigma_1^2, 0.3*sigma_1*sigma_2,
                  0.3*sigma_0*sigma_2, 0.3*sigma_1*sigma_2, sigma_2^2), nrow = 3)
  b <- MASS::mvrnorm(n, c(0.1, 0.2, 0.3), Sigma)
  b0 <- b[,1]
  b1 <- b[,2]
  b2 <- b[,3]
  B0 <- b0[ID]; B1 <- b1[ID]; B2 <- b2[ID]
  
  #b-spline
  bs_m <- bs(time, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)
  
  #fix effect coef
  epsilon <- rnorm(n*obs, 0, 4)
  
  y <- beta*group + B0*bs_m[,1] + B1*bs_m[,2] + B2*bs_m[,3] + epsilon
  df <- data.frame(ID = ID, time = time, group = group, y = y, bs1 = bs_m[,1], bs2 = bs_m[,2], bs3 = bs_m[,3])
  
  lme.bs <- lmer(y ~ -1 + group + (-1 + bs1 + bs2 + bs3 | ID), data = df, control=lmerControl(check.conv.grad= .makeCC("ignore",tol = 2e-3, relTol = NULL)))
  
  #coef
  coef_est <- as.numeric(fixef(lme.bs))
  #variance-covariance matrix
  cov_m <- as.matrix(Matrix::bdiag(VarCorr(lme.bs)))
  sigma1_est <- sqrt(cov_m[1,1])
  sigma2_est <- sqrt(cov_m[2,2])
  sigma3_est <- sqrt(cov_m[3,3])
  cov12_est <- cov_m[1,2]
  cov13_est <- cov_m[1,3]
  cov23_est <- cov_m[2,3]
  return(c(coef_est, sigma1_est, sigma2_est, sigma3_est, cov12_est, cov13_est, cov23_est))
}
```

parallel computing
```{r}
library(doParallel)
n_core <- detectCores() - 1
cl <- parallel::makeCluster(n_core)
doParallel::registerDoParallel(cl)
```

data generation (n=10000,obs=40)
```{r}
start.time <- Sys.time()
sim_n <- 1000
# Using foreach to replicate the function in parallel
results <- foreach(i = 1:sim_n, .packages=c("lme4", "splines")) %dopar% {
  simulation(n = 10000, obs = 40, seed = i) #n=10000,5000,2000,500,100,50; obs=40,10,5
}
end.time <- Sys.time()
end.time - start.time

#bias
mean(sapply(results, function(x) x[1] - beta)); mean(sapply(results, function(x) x[2] - sigma_0)); mean(sapply(results, function(x) x[3] - sigma_1)); mean(sapply(results, function(x) x[4] - sigma_2)); mean(sapply(results, function(x) x[5] - 0.3*sigma_0*sigma_1)); mean(sapply(results, function(x) x[6] - 0.3*sigma_0*sigma_2)); mean(sapply(results, function(x) x[7] - 0.3*sigma_1*sigma_2))

#MSE
mean(sapply(results, function(x) (x[1] - beta)^2)); mean(sapply(results, function(x) (x[2] - sigma_0)^2)); mean(sapply(results, function(x) (x[3] - sigma_1)^2)); mean(sapply(results, function(x) (x[4] - sigma_2)^2));
mean(sapply(results, function(x) (x[5] - 0.3*sigma_0*sigma_1)^2)); mean(sapply(results, function(x) (x[6] - 0.3*sigma_0*sigma_2)^2)); mean(sapply(results, function(x) (x[7] - 0.3*sigma_1*sigma_2)^2))
```

boxplot
```{r}
library(ggplot2)
library(reshape2)
beta_bias <- sapply(results, function(x) x[1] - beta)
sigma1 <- sapply(results, function(x) x[2] - sigma_0)
sigma2 <- sapply(results, function(x) x[3] - sigma_1)
sigma3 <- sapply(results, function(x) x[4] - sigma_2)
bias <- data.frame(beta_bias, sigma1, sigma2, sigma3)
ggplot(melt(bias), aes(x=variable, y=value)) +
  geom_boxplot(aes(fill=variable)) + 
  ylim(-1.5, 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic()
```


### 4.4 Comparison of different methods
generate data
```{r}
set.seed(1234)
############ group 1 ############
n <- 2500; obs <- 40
beta <- 1.5
time <- rweibull(n*obs, 2)
ID <- rep(1:n, obs)
group <- rep(1, n)
group <- group[ID]

#generate random effect coefficients
sigma_0 <- 1.5; sigma_1 <- 1.0; sigma_2 <- 2
Sigma <- matrix(c(sigma_0^2, 0.3*sigma_0*sigma_1, 0.3*sigma_0*sigma_2,
                  0.3*sigma_0*sigma_1, sigma_1^2, 0.3*sigma_1*sigma_2,
                  0.3*sigma_0*sigma_2, 0.3*sigma_1*sigma_2, sigma_2^2), nrow = 3)
b <- MASS::mvrnorm(n, c(0, 0, 0), Sigma)
b0 <- b[,1]
b1 <- b[,2]
b2 <- b[,3]
B0 <- b0[ID]; B1 <- b1[ID]; B2 <- b2[ID]
mean(b0); mean(b1); mean(b2); cov(b)

epsilon <- rnorm(n*obs, 0, 4)

#b-spline
degree <- 3
bs_m <- bs(time, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)

y <-  B0*bs_m[,1] + B1*bs_m[,2] + B2*bs_m[,3] + epsilon 
df <- data.frame(ID = ID, time = time, group = group, y = y)

############ group 2 ############
set.seed(1235)
n <- 2500; obs <- 40
time <- rweibull(n*obs, 2)
ID <- rep((n+1):(2*n), obs)
group <- rep(0, n)
group <- group[ID-n]

epsilon <- rnorm(n*obs, -0.03, 4.5)

y <-  epsilon 
df1 <- data.frame(ID = ID, time = time, group = group, y = y)
df <- rbind(df, df1)

# fit the model
lme.bs <- lmer(y ~ -1 + group + (-1 + bs(time, Boundary.knots = quantile(time, c(0,0.75)), degree = degree)| ID), data = df, control=lmerControl(check.conv.grad= .makeCC("ignore",tol = 2e-3, relTol = NULL)))
summary(lme.bs)
```

```{r, warning=FALSE}
#predictive data
long_time <- c()
ID <- c()
group <- c()
for (id in unique(df$ID)) {
  long_time <- append(long_time, seq(min(df$time[df$ID == id]), max(df$time[df$ID == id]), by = max(df$time)/1000))
  ID <- append(ID, rep(id, length(seq(min(df$time[df$ID == id]), max(df$time[df$ID == id]), by = max(df$time)/1000))))
  group <- append(group, rep(df$group[df$ID == id][1], length(seq(min(df$time[df$ID == id]), max(df$time[df$ID == id]), by = max(df$time)/1000))))
}

new_sim <- data.frame(ID = ID, time = long_time, group = group)
new_sim$pred <- predict(lme.bs, new_sim)

#cumulative variation
degree <- 3;interval <- 0.001;fixcoef <- fixef(lme.bs); rancoef <- ranef(lme.bs)$ID
arc <- c()
for (id in unique(new_sim$ID)) {
  endtime <- max(new_sim$time[new_sim$ID == id])
  starttime <- min(new_sim$time[new_sim$ID == id])
  group_index <- new_sim$group[new_sim$ID == id][1]
  b_vec <- unlist(c(fixcoef, rancoef[rownames(rancoef) == id,]))
  gt <- seq(starttime, endtime, by = interval)
  b_mat <- cbind(group_index, bs(gt, Boundary.knots = quantile(time, c(0,0.75)), degree = degree))
  g_grid <- b_mat %*% b_vec
  arc <- append(arc, sum(sqrt((g_grid[-length(gt)]-g_grid[-1])^2+interval^2)))
}

```

Kruskal-Wallis test
```{r}
#subject-level measurements
test_df <- df %>% group_by(ID) %>%
  summarise(group = group[1], mean = mean(y), sd = sd(y), cv = mean(y)/sd(y)) 
test_df$cum_v <- arc
test_df$group <- as.factor(test_df$group)
test_df <- as.data.frame(test_df)

#Kruskal-Wallis test
krus_test_comp_5000 <- c()
for (i in 3:6) {
  krus_test <- kruskal.test(test_df[, i], test_df$group)$p.value
  krus_test_comp_5000 <- c(krus_test_comp_5000, krus_test)
}
krus_test_comp_5000
```

